Linking assumptions and number of response options modulate inferred scalar implicature rate

### 2018-05-09 writeup.Rmd
minor proofreading edits

Our primary goal in this study was to test whether the estimated implicature rate in the experimental task is affected by the linking assumption and the number of response options available to participants. To this end, we only analyzed the critical trials (exhaustive and scalar). In particular, we classified each data point from critical trials as constituting an implicature (1) or not (0) under the strong and Weak link. Figure \@ref(fig:implicatureRatePlot) shows the resulting implicature rates by condition and link.

@@

The model provided evidence for the following effects: First, there was a main effect of trial type such that scalar trials resulted in greater implicature rates than exhaustive trials (Mean Estimate = `r round(fixed_effects$Estimate[6],2)`, 95% Credible Interval=[`r round(fixed_effects$X2.5.ile[6],2)`, `r round(fixed_effects$X97.5.ile[6],2)`]). Second, there was an interaction between link and number of response options such that the quaternary task (Mean Estimate = `r round(fixed_effects$Estimate[7],2)`, 95% Credible Interval=[`r round(fixed_effects$X2.5.ile[7],2)`, `r round(fixed_effects$X97.5.ile[7],2)`]) and the quinary task  (Mean Estimate = `r round(fixed_effects$Estimate[8],2)`, 95% Credible Interval=[`r round(fixed_effects$X2.5.ile[8],2)`, `r round(fixed_effects$X97.5.ile[8],2)`]) with a weak link resulted in greater implicature rates. Finally, there was a three-way interaction between link, trial type, and number of response options (Mean Estimate = `r round(fixed_effects$Estimate[15],2)`, 95% Credible Interval=[`r round(fixed_effects$X2.5.ile[15],2)`, `r round(fixed_effects$X97.5.ile[15],2)`]). One interpretation of this interaction is that the difference between the weak and strong link on scalar trials in the quinary task was smaller than on exhaustive trials, though we believe this is not too interesting, given that the binary reference level implicature estimate was lower for exhaustive trials in the first place. Crucially, both number of response options and link affect the inferred implicature rate.

@@

While the binary truth value judgement task avoids the analytic decision between strong and Weak linking assumptions, the results reported here suggest that binary tasks can also underestimate participants' pragmatic competence. In binary tasks, participants are often given the lowest and highest points on a scale ("wrong" vs. "right") and are asked to report pragmatic infelicities using the lowest point (e.g. "wrong"). The study reported here showed that on trials with true but pragmatically infelicitous descriptions, participants often avoided the lowest point on the scale if they were given more intermediate options. Even though the option "wrong" was available to participants in all tasks, participants in tasks with intermediate options chose it less often. In computing implicature rate, this pattern manifested itself as a decrease in implicature rate under the Strong link when more response options were provided, and an increase in implicature rate under the Weak link when more response options were provided. These conclusions are in line with @Katsos2011's argument that pragmatic violations are not as severe as semantic violations and participants do not penalize them as much. Providing participants with only the extreme ends of the scale (e.g. wrong/right, false/true) when pragmatic violations are considered to be of an intermediate nature risks misrepresentation of participants' pragmatic competence. It further suggests that in studies that use binary tasks to investigate response-contingent processing, proportions of "literal" responses may be a composite of both literal and pragmatic underlying interpretations that just happen to get mapped differently onto different response options by participants.

@@

**The following is all new** In an RSA model, $S_1$(u | s) will be defined for any possible combination of utterance and world state. One upshot of this is that for the purposes of our linking hypothesis, participants are modeled as employing the same heuristic - does $S_1$(u | s) exceed the threshold? - in both 'implicature' and 'non-implicature' conditions of a truth value judgment task experiment. That is, participants never evaluate utterances directly on the basis of logical truth or falsity: for example, our blindfolded character Bob's guess of "cat and dog" on a cat and dog card trial is "right" to the vast majority of participants not because the guess is logically true but because $S_1$("cat and dog" | cat and dog) is exceedingly high.

@@

For the time being, however, we present the above analysis as a proof of concept for the following idea: by relaxing the assumptions of the traditional view of scalar implicature (namely, that scalar implicatures either are or are not calculated, and that behavior on sentence verification tasks directly reflects this binary interpretation process), we can propose quantitative models of the variation in behavior we observe in experimental settings. We note that the linking analysis proposed here is just one in the space of possible analyses when traditional assumptions about scalar implicature are relaxed. For example, one might reject this threshold-based analysis in favor of one whereby responses are the outcomes of sampling on the (pragmatic speaker or pragmatic listener) probability distributions provided by an RSA model. We must leave this investigation to future work, but for the time being we emphasize that such quantitative, data-driven and systematic model criticism/comparison is made available to researchers in experimental pragmatics by revising core assumptions about the nature of scalar implicature. Though we no longer have a crisp notion of scalar implicature as something that is or is not 'calculated' in interpretation, we have new flexibility to explicitly discuss categorical behavior in experimental settings.

>>>>

Our primary goal in this study was to test whether the estimated implicature rate in the experimental task is affected by the linking assumption and the number of response options available to participants. To this end, we only analyzed the critical trials (exhaustive and scalar). In particular, we classified each data point from critical trials as constituting an implicature (1) or not (0) under the Strong and Weak link. Figure \@ref(fig:implicatureRatePlot) shows the resulting implicature rates by condition and link.

@@

The model provided evidence for the following effects: First, there was a main effect of trial type such that scalar trials resulted in greater implicature rates than exhaustive trials (Mean Estimate = `r round(fixed_effects$Estimate[6],2)`, 95% Credible Interval=[`r round(fixed_effects$X2.5.ile[6],2)`, `r round(fixed_effects$X97.5.ile[6],2)`]). Second, there was an interaction between link and number of response options such that the quaternary task (Mean Estimate = `r round(fixed_effects$Estimate[7],2)`, 95% Credible Interval=[`r round(fixed_effects$X2.5.ile[7],2)`, `r round(fixed_effects$X97.5.ile[7],2)`]) and the quinary task  (Mean Estimate = `r round(fixed_effects$Estimate[8],2)`, 95% Credible Interval=[`r round(fixed_effects$X2.5.ile[8],2)`, `r round(fixed_effects$X97.5.ile[8],2)`]) with a weak link resulted in greater implicature rates. Finally, there was a three-way interaction between link, trial type, and number of response options (Mean Estimate = `r round(fixed_effects$Estimate[15],2)`, 95% Credible Interval=[`r round(fixed_effects$X2.5.ile[15],2)`, `r round(fixed_effects$X97.5.ile[15],2)`]). One interpretation of this interaction is that the difference between the Weak and Strong link on scalar trials in the quinary task was smaller than on exhaustive trials, though we believe this is not too interesting, given that the binary reference level implicature estimate was lower for exhaustive trials in the first place. Crucially, both number of response options and link affect the inferred implicature rate.

@@

While the binary truth value judgement task avoids the analytic decision between Strong and Weak linking assumptions, the results reported here suggest that binary tasks can also underestimate participants' pragmatic competence. In binary tasks, participants are often given the lowest and highest points on a scale ("wrong" vs. "right") and are asked to report pragmatic infelicities using the lowest point (e.g. "wrong"). The study reported here showed that on trials with true but pragmatically infelicitous descriptions, participants often avoided the lowest point on the scale if they were given more intermediate options. Even though the option "wrong" was available to participants in all tasks, participants in tasks with intermediate options chose it less often. In computing implicature rate, this pattern manifested itself as a decrease in implicature rate under the Strong link when more response options were provided, and an increase in implicature rate under the Weak link when more response options were provided. These conclusions are in line with @Katsos2011's argument that pragmatic violations are not as severe as semantic violations and participants do not penalize them as much. Providing participants with only the extreme ends of the scale (e.g. wrong/right, false/true) when pragmatic violations are considered to be of an intermediate nature risks misrepresentation of participants' pragmatic competence. It further suggests that in studies that use binary tasks to investigate response-contingent processing, proportions of "literal" responses may be a composite of both literal and pragmatic underlying interpretations that just happen to get mapped differently onto different response options by participants.

@@

**The following is all new** In an RSA model, $S_1$(u | s) will be defined for any possible combination of possible utterance and possible world state. One upshot of this is that for the purposes of our linking hypothesis, participants are modeled as employing the same heuristic - does $S_1$(u | s) exceed the threshold? - in both 'implicature' and 'non-implicature' conditions of a truth value judgment task experiment. That is, participants never evaluate utterances directly on the basis of logical truth or falsity: for example, our blindfolded character Bob's guess of "cat and dog" on a cat and dog card trial is "right" to the vast majority of participants not because the guess is logically true but because $S_1$("cat and dog" | cat and dog) is exceedingly high.

@@

For the time being, however, we present the above analysis as a proof of concept for the following idea: by relaxing the assumptions of the traditional view of scalar implicature (namely, that scalar implicatures either are or are not calculated, and that behavior on sentence verification tasks directly reflects this binary interpretation process), we can propose quantitative models of the variation in behavior we observe in experimental settings. We note that the linking analysis proposed here is just one in the space of possible analyses when traditional assumptions about scalar implicature are relaxed. For example, one might reject this threshold-based analysis in favor of one whereby responses are the outcomes of sampling on the (pragmatic speaker or pragmatic listener) probability distributions provided by an RSA model. We must leave this investigation to future work, but for now we emphasize that this kind of quantitative, data-driven and systematic model criticism is made available to researchers in experimental pragmatics by revising core assumptions about the nature of scalar implicature. Though we no longer have a crisp notion of scalar implicature as something that is or is not 'calculated' in interpretation, we have new flexibility to explicitly discuss categorical behavior in experimental settings.
